Genome Function Phylogenetics

### 2019-04-16 final_project/paper/Paper.Rmd
Rephrased methods part

---
title: "Gene Function Phylogenetics"
author: "Dennis Psaroudakis"
date: "April 18th 2019"
keep_md: yes
output:
  tufte::tufte_handout:
    latex_engine: xelatex
    citation_package: natbib
  tufte::tufte_html: default
bibliography: resources/bib.bib
link-citations: yes
classoption: justified
language: en-US
header-includes:
  - \usepackage[osf]{libertine}
  - \citestyle{plain}
---

@@

`r newthought("Introduction, Data...")`

## Method
Starting point of our method is a functional annotation set for each genome $S_a$, consisting of tuples $(G_i, T_i)$ with $G_i \in$ Genes in genome $S_a$ and $T_i \in$ Terms in the Gene Ontology. So the annotation set maps each gene of the genome to $n$ GO terms.
\marginnote[-4em]{
\texttt{Gene\ \ \ \ \ \ \ \ \ GO Term\\
Os01g0601625	GO:0050896\\
Os01g0601625	GO:0016021\\
Os01g0601625	GO:0016301\\
Os01g0601651	GO:0003677\\
Os01g0601651	GO:0009699\\
Os01g0601651	GO:0050790\\
Os01g0601651	GO:0050794\\
Os01g0601651	GO:0050896\\
Os01g0601675	GO:0007275\\
Os01g0601675	GO:0016310\\
Os01g0601675	GO:0050789\\
...
}
}

Now we use the structure of the Gene Ontology to obtain all ancestors $A_i$ of GO term $T_i$ in the GO hierarchy and combine all these annotations and their ancestors, irrespective of the Gene they are associated with: $\bigcup_{i=1}^{x}(T_i \cup A_{i})$ The resulting set is a set of all the GO terms present in the genome annotation and their ancestors.

After this is done for all annotation sets, we introduce the Jaccard distance as a metric of (dis-)similarity between the annotation sets:

@@

\begin{marginfigure}
\includegraphics{resources/hF-graphical.png}
\caption{\textbf{(Zoom in)}. Graphic of how we use the Gene Ontology tree to calculate the overlap (= intersection) of two GO terms (the ones marked with a leaf)}
\label{fig:marburg_virus}
\end{marginfigure}

Then finally I can apply Neighbor-Joining to the distance matrix I generated this way.

>>>>

---
title: "Genome Function Phylogenetics"
author: "Dennis Psaroudakis"
date: "April 18th 2019"
keep_md: yes
output:
  tufte::tufte_handout:
    latex_engine: xelatex
    citation_package: natbib
  tufte::tufte_html: default
bibliography: resources/bib.bib
link-citations: yes
classoption: justified
language: en-US
header-includes:
  - \usepackage[osf]{libertine}
  - \citestyle{plain}
---

@@

`r newthought("Introduction, Gene Ontology, Data...")`

## Method
Starting point of our method are the functional annotation sets, one for each genome, which annotate every gene in the genome with one or more GO terms.
In more mathematical terms the genome annotation set is a list of tuples $(G, T)$ with $G \in$ Genes in that genome and $T \in$ Terms in the Gene Ontology.
\marginnote[-5em]{
\texttt{Gene\ \ \ \ \ \ \ \ \ GO Term\\
Os01g0601625	GO:0050896\\
Os01g0601625	GO:0016021\\
Os01g0601625	GO:0016301\\
Os01g0601651	GO:0003677\\
Os01g0601651	GO:0009699\\
Os01g0601651	GO:0050790\\
Os01g0601651	GO:0050794\\
Os01g0601651	GO:0050896\\
Os01g0601675	GO:0007275\\
Os01g0601675	GO:0016310\\
Os01g0601675	GO:0050789\\
...
}
}

We can use the hierarchical structure of the Gene Ontology to obtain the ancestors $A_i$ of any term $T_i$; in other words the gene $G_i$ is not just annotated with the term $T_i$ itself but also with all GO terms that are a more general statement of that term (e.g. any gene that is part of a metabolic process is thereby also part of a biological process).
We do that for all terms $T$ in the dataset and combine all of the terms and their ancestors into one big genome-wide set $S$, irrespectively of the gene they were originally associated with:  $S = \bigcup_{i=1}^{x}(T_i \cup A_{i})$.

When this superset of annotations is created for each of the datasets, we can use the Jaccard Distance as a measure of how (dis-)similar any two sets are from each other, or in biological terms how different the two genomes are on a functional level:

@@

\begin{marginfigure}
The general idea of using the Jaccard Distance in this context is to measure the overlap of two subtrees in the GO hierarchy.
Say, for simplicity, that we're looking at two genomes (here called Gold Standard and Prediction) that each only contain one single GO term (marked by a leaf).
First, we add all ancestors of that leaf term to each subtree.
Then, we determine the overlap (which corresponds to $S_a \cap S_b$), and divide the number of nodes in this overlap by the number of nodes in either of the two subtrees ($S_a \cup S_b$).
\vspace{0.2em}

\includegraphics{resources/hF-graphical.png}

\vspace{0.5em}
\noindent In the case of this example, the Jaccard Distance of Gold Standard and Prediction would be $1 - \frac{4}{9} = \frac{5}{9}$
\end{marginfigure}

Applying this formula to all pairwise combinations of the genomes we're looking at yields a $S \times S$ distance matrix that can then serve as the input for a neighbor joining algorithm.
